Reparametrization of COM-Poisson Regression Models with Applications in the Analysis of Experimental Data

### 2018-01-27 article.Rnw
Improve text by Hinde's suggestions and some corrections

% -------------------------------------------
% Abstract
\Abstract{ In the analysis of count data often the equidispersion
  assumption is not suitable, hence the Poisson regression model is
  inappropriate.  As a generalization of the Poisson distribution the
  COM-Poisson distribution can deal with under-, equi- and overdispersed
  count data.  It is a member of the exponential family of distributions
  and has the Poisson and geometric distributions as special cases as
  well as the Bernoulli distribution as a limiting case.  In spite of
  the nice properties of the COM-Poisson distribution, its location
  parameter does not correspond to the expectation, which in turn
  becomes challenging the interpretation of the regression models
  specified using such a distribution. In this paper, we propose a
  straightforward reparametrization of the COM-Poisson distribution
  based on an approximation to the expectation of this distribution. The
  main advantage of our new parametrization is the straightforward
  interpretation of the regression coefficients in terms of the
  expectation of the count response variable, as usual in the context of
  generalized linear models. Furthermore, the estimation and inference
  for the new COM-Poisson regression model can be done based on the
  likelihood paradigm. We carried out simulation studies to verify the
  finite sample properties of the maximum likelihood estimators. The
  results from our simulation study show that the maximum likelihood
  estimators are unbiased and consistent for both regression and
  dispersion parameters. We observed that the empirical correlation
  between the regression and dispersion parameter estimators is close to
  zero, which suggests that these parameters are orthogonal. We
  illustrate the application of the proposed model through the analysis
  of three data sets with over-, under- and equidispersed count data.
  The study of distribution properties through a consideration of
  dispersion, zero-inflated and heavy tail indices, and the results of
  data analysis show the flexibility over standard
  approaches. Therefore, we encourage the application of the new
  parametrization for the analysis of count data in the context of
  COM-Poisson regression models.  The computational routines for fitting
  the original and new version of the COM-Poisson regression model and
  the analyzed data sets are available in the supplementary material.
}

@@

Count data are random variables that assume non-negative integer values
and represent the number of times an event occurs in the observation
unit. This kind of data is common in crop sciences such as the number of
grains produced by a plant, number of fruits produced by a tree, number
of insects captured by a trap, to cite but a few.  Since the seminal
paper of \citet{Nelder1972} where the class of the generalized linear
models (GLMs) was introduced, the analysis of count data often employs
the Poisson regression model. This model provides a suitable strategy
for the analysis of count data and an efficient Newton scoring algorithm
can be used for fitting the model.

In spite of the advantages of the Poisson regression model, the Poisson
distribution has only one parameter, which represents both the
expectation and variance of the count random variable.  This restriction
in the relationship between the expectation and variance induced by the
Poisson distribution is referred as equidispersion.  However, in
practical data analysis such a restriction can be unsuitable, since the
observed data can present both the variance smaller or larger than the
expectation, leading to the cases of under and overdispersion,
respectively.  The main problem of the application of the Poisson
regression model to non equidispersed count data is that the standard
errors associated with the regression coefficients are inconsistently
estimated, which in turn can lead to misleading
inferences~\citep{Winkelmann1995, Bonat2017}.

In practice, overdispersion is largely reported in the literature and
may occur due to the absence of relevant covariates, heterogeneity of
sampling units, different domain size not considered in the analysis and
excess of zeros~\citep{Hinde1998}.  The case of underdispersion is less
report in the literature, however, it has been of increasing interest in
the statistical community. The processes that reduce the variability are
not as well-known as those leading to extra variability.  For the same
reason, there are a few approaches to deal with underdispersed count
data. The explanatory mechanisms leading to underdispersion may be
related to the underline Poisson process generating the count data.
When the time between events is not exponentially distributed, the number
of events can be over or underdispersed, this process motivated the
class of duration dependence models~\citep{Winkelmann1995}.  Another
possibility to explain the underdispersion phenomenon is order
statistics.  Suppose $Y_i$ is Poisson distributed then the maximum of
$Y_i$ is underdispersed~\citep{Steutel1989}.

The strategies for constructing alternative count distributions are
related with the causes of the non-equidispersion. In special to the
overdispersion case the Poisson mixture models are widely applied.  One
popular example of this approach is the negative-binomial model, where
the expectation of the Poisson distribution is assumed gamma
distributed.  However, other distributions can be used to represent the
random effects.  For example the Poisson-Tweedie model~\citep{Bonat2017}
and its special cases as the Poisson inverse-Gaussian and Neyman-Type A
consider that the random effects are Tweedie, inverse Gaussian and
Poisson distributed, respectively.  The Gamma-Count distribution assumes
a gamma distribution for the time between events, thus it can deal with
underdispersed as well as overdispersed count
data~\citep{Zeviani2014}. The COM-Poisson distribution is obtained by a
generalization of the Poisson distribution allowing for a non-linear
decrease in the ratios of successive probabilities~\citep{Shmueli2005}.

The COM-Poisson distribution is a member of the exponential family and
it has the Poisson and geometric distributions as special cases as well
as the Bernoulli distribution as a limiting case. It can deal with both
under and overdispersed count data.  Some recently applications of the
COM-Poisson distribution include \citet{Lord2010} for the analysis of
traffic crash data, \citet{Sellers2010} for the modelling of airfreight
breakage and book purchases and \citet{Huang2017} to the analysis of
attendance data, takeover birds and cotton bolls count.  The main
disadvantage of the COM-Poisson regression model as presented
in~\citet{Sellers2010} is that its location parameter does not
correspond to the expectation of the distribution, which in turn becomes
the interpretation of the regression coefficients challenging and no
comparable with standard approaches as the Poisson and negative binomial
regression models. \citet{Huang2017} proposed a mean-parametrization of
the COM-Poisson distribution in order to avoid such an issue.  In this
approach the mean parameter is obtained by solving an non-linear
equation defined as an infinite sum.  Consequently, it is
computationally demanding and liable to numerical problems.

The main goal of this article is to propose a novel COM-Poisson
parametrization based on the mean approximation presented by
\citet{Shmueli2005}. In this parametrization, the probability mass
function is written in terms of $\mu$ and $\phi$, where $\mu$ is the
expectation and $\phi$ is the dispersion parameter. In contrast to the
original parametrization, the proposed parametrization leads to
regression coefficients directly associated with the expectation of the
response variable as usual in the context of generalized linear models.
Consequently, the obtained regression coefficients are comparable with
the ones obtained by standard approaches as the Poisson and negative
binomial regression models.  Furthermore, our novel COM-Poisson
parametrization is simpler than the strategy proposed
by~\cite{Huang2017}, since it does not involve any numerical method for
solving non-linear equations.

This paper is organized as follows. In \autoref{background} we present
the COM-Poisson distribution and the strategy proposed
by~\cite{Huang2017}.  The proposed reparametrization, assessment of
moment approximations, and study of distribution properties are
considered in the~\autoref{reparametrization}. In
the~\autoref{estimation-and-inference} we present the estimation and
inference for the novel COM-Poisson regression model based on the
likelihood paradigm. The properties of the maximum likelihood estimators
are assess in~\autoref{simulation-study} through simulation studies.  We
illustrate the application of the new COM-Poisson regression model
through the analysis of three data sets. We provide an \texttt{R}
implementation of the COM-Poisson and reparameterized COM-Poisson
regression models as well as the analyzed data sets in the supplementary
material.\footnote{Available on
  \texttt{\url{http://www.leg.ufpr.br/~eduardojr/papercompanions}}
  \label{papercompanion}.}.

@@

The COM-Poisson distribution generalizes the Poisson distribution in
terms of the ratio between the probabilities of two consecutive events
by adding an extra dispersion parameter~\citep{Sellers2010}.  Let $Y$ be
a COM-Poisson random variable, then
$$\frac{\Pr(Y=y-1)}{\Pr(Y=y)} = \frac{y^\nu}{\lambda}$$ while for the
Poisson distribution this ratio is $\frac{y}{\lambda}$. It allows the
COM-Poisson distribution deals with non equidispersed count data.  The
probability mass function of the COM-Poisson distribution is given by
\begin{equation}
  \label{eqn:pmf-cmp}
  \Pr(Y=y \mid \lambda, \nu) = \frac{\lambda^y}{(y!)^\nu Z(\lambda, \nu)},
  \qquad y = 0, 1, 2, \ldots,
\end{equation}
where $\lambda > 0$, $\nu \geq 0$ and $Z(\lambda, \nu) =
\sum_{j=0}^\infty \frac{\lambda^j}{(j!)^\nu}$ is a normalizing
constant that depends on both parameters.

The $Z(\lambda, \nu)$ series diverges theoretically only when $\nu=0$
and $\lambda \geq 1$, but numerically for small values of $\nu$ combined
with large of $\lambda$, the sum is so huge it causes
overflow. \autoref{tab:convergenceZ} shows the values of the
normalizing constants using one thousand increments, that is,
$\sum_{j=0}^{1000}\lambda^j/(j!)^\nu$ for different values of $\lambda$
and $\phi$.

@@

In the first line of \autoref{tab:convergenceZ} we have mathematically
divergent series, because $\sum_{j=0}^\infty\lambda^j$, when
$\lambda \geq 1$, is divergent.  In other cases the series diverges
numerically, due to the computational storage limitation.  For both
divergences it is impossible to compute probabilities, therefore, this
acts as a restriction in the parameter space.

@@

The COM-Poisson regression model was proposed by~\citet{Sellers2010},
using the original parametrization. In this case, the COM-Poisson
regression model is $\log(\lambda_i) = \bm{x}_i^\top \bm{\beta}$ and the
relationship between E$(Y_i)$ and $\bm{x}_i$ is modelled indirectly.
\citet{Huang2017} shows how to model directly the expectation of the
COM-Poisson distribution in a suitable parametrization. In the
\autoref{eqn:pmf-cmp}, \Citeauthor{Huang2017} proposes that the
parameter $\lambda$ as a function of $\mu$ and $\nu$, given by the
solution to
\begin{equation*}
  \sum_{j=0}^{\infty} (j - \mu) \frac{\lambda^j}{(y!)^\nu} = 0\,.
\end{equation*}
Thus the mean-parametrized COM-Poisson regression model is $\log(\mu_i)
= \bm{x}_i^\top \bm{\beta}$. In this article, we propose an alternative
mean-parametrization of the COM-Poisson distribution in order to avoid the
limitations of the original parametrization and the numerical complexity
of the \Citeauthor{Huang2017}'s approach.

@@

The proposed reparametrization of COM-Poisson models is based on
mean approximation (\autoref{eqn:mean-approx}). We introduced new
parameter $\mu$, using this approximation,
\begin{equation}
  \label{eqn:repar-cmp}
  \mu = h(\lambda, \nu) = \lambda^{1/\nu} - \frac{\nu - 1}{2\nu}
  \quad \Rightarrow \quad
  \lambda = h^{-1}(\mu, \nu) = \left (\mu +
    \frac{(\nu - 1)}{2\nu} \right )^\nu.
\end{equation}
The dispersion parameter is taken on the log scale for computational
convenience, thus $\phi = \log(\nu)$, $\phi \in \mathbb{R}$. The
interpretation of $\phi$ is the same as the $\nu$, but in another
scale. For $\phi < 0$ and $\phi > 0$ we have the overdispersed and
underdispersed cases, respectively.  When $\phi=0$ we have Poisson
distribution as the special case.

@@

In order to assess the accuracy of the moments approximations
(\autoref{eqn:mean-approx}), \autoref{fig:approx-plot} presents the
quadratic errors for the expectation (a) and variance (b).  The
quadratic errors were obtained by $[\mu - \text{E}(Y)]^2$ for the
expectation and by $[ \mu \exp(-\phi) - \text{Var}(Y)]^2$ for the
variance. In both cases $\text{E}(Y)$ and Var$(Y)$ were computed
numerically. The dotted lines represent the border between the regions
$\nu \leq 1$ and $\lambda > 10^\nu$, in the $\mu$ and $\phi$ scale.

<<approx-plot, fig.height=3, fig.width=6, out.width=".8\\textwidth", fig.cap="Quadratic errors for the approximation of the expectation (a) and variance (b). Dotted lines represent the restriction for suitable approximations given by \\cite{Shmueli2005}.">>=

@@

The results in \autoref{fig:approx-plot} show that the mean
approximation is accurate, the largest quadratic error is
\Sexpr{max(grid[["emu"]])} for the parameter values evaluated. For the
variance approximation, the largest quadratic error was
\Sexpr{max(grid[["eva"]])} and it occurs for negative values of
$\phi$. Interestingly, the errors are larger to negative values of
$\phi$ and present no clear relation with $\mu$, opposite to the regions
gives by \citet{Shmueli2005} ($\phi \leq 0$ and
$\mu > 10 - \frac{\exp(\phi) - 1}{2\exp(\phi)}$).

@@

These indexes are discussed by \citet{Bonat2017} to study the
flexibility of Poisson-Tweedie distribution, and \citet{Puig2006} to
describe count distributions. Regarding the COM-Poisson$_\mu$
distribution, we present in \autoref{fig:indexes-plot} the relationship
between mean and variance (a), the dispersion and zero-inflation indexes
(b--c) for different values of $\mu$ and $\phi$ and heavy-tail index
(d) for $\mu=25$ and different values of $y$ and $\phi$.

@@

<<indexes-plot, fig.width=9.5, fig.height=4, fig.cap="Indexes for COM-Poisson distribution. Mean and variance relationship (a), dispersion, zero-inflation and heavy-tail indexes (b--d) for different parameter values. Dotted lines represents the Poisson special case.">>=

@@

\autoref{fig:indexes-plot} shows that the indexes are slightly depend on
the expected values and tend to stabilize for large values of
$\mu$. Consequently, the mean and variance relationship
\autoref{fig:indexes-plot}(a) is proportional to the dispersion
parameter $\phi$. In terms of moments, this leads to a specification
indistinguishable from the quasi-Poisson regression model. The
dispersion indexes in \autoref{fig:indexes-plot}(b) show that the
distribution is suitable to deal to dispersed counts, of course. For the
parameter values evaluated the largest DI was \Sexpr{madi} and smallest
was \Sexpr{midi}. \autoref{fig:indexes-plot}(c) shows the COM-Poisson
can handle with a limited amount of zero-inflation, in cases of
overdispersion ($\phi < 0$). On the other hand, for $\phi > 0$
(underdispersion) this distribution is suitable to deal with
zero-deflated counts. Heavy-tail indexes in
\autoref{fig:indexes-plot}(d) indicates the distribution is in general
light tailed distribution, i.e. $HT \to 0$ for $y \to \infty$.

@@

In this section we shall describe the estimation and inference to the
two forms of the COM-Poisson regression model based on the maximum
likelihood method. The log-likelihood function for a set of independent
observations $y_i$, $i=1,2,\ldots,n$ from the COM-Poisson$_\mu$
distribution has the following form,
\begin{equation}
  \label{eqn:ll-rcmp}
  \ell = \ell(\bm{\beta}, \phi \mid \bm{y}) =
  e^\phi \left [
    \sum_{i=1}^n y_i
    \log \left( \mu_i + \frac{e^\phi-1}{2e^\phi} \right ) -
    \sum_{i=1}^n \log(y_i!) \right ] -
  \sum_{i=1}^n \log(Z(\mu_i, \phi)),
\end{equation}
where $\mu_i = \exp(\bm{x}_i^\top\bm{\beta})$, with
$\bm{x}_i^\top = (x_{i1},\, x_{i2},\, \ldots,\, x_{ip})$ is a vector of
known covariates for the $i$-th observation, and $(\bm{\beta},\, \phi)
\in \mathbb{R}^{p+1}$. The normalizing constant $Z(\mu_i, \phi)$ is
given by
\begin{equation}
\label{eqn:infseries}
  Z(\mu_i, \phi) = \sum_{j=0}^\infty \left [ \left (
    \mu_i + \frac{e^\phi - 1}{2e^\phi} \right )^{je^\phi}
  \frac{1}{(j!)^{e^\phi}} \right ].
\end{equation}
The evaluation of the log-likelihood function for each observation
involves the computation of the infinite series
(\autoref{eqn:infseries}).  Thus, the fitting procedure is computational
expensive for regions of the parameter space where the convergence of
the infinite sum is slow.

@@

In the applications we fitted the quasi-Poisson model
\citep{Wedderburn1974} as a baseline model. This approach is based on
second-moment assumption that allows more flexibility to model. In this
case the variance of the response variable is fixed by an additional
parameter $\sigma$, $\textrm{Var}(Y_i)=\sigma \mu_i$. These models are
fitted in software \texttt{R} using the function \texttt{glm(..., family
  = quasipoisson)}.

@@

We considered average counts varying from $3$ to $27$ according to a
regression model with a continuous and a categorical covariate. The
continuous covariate~($\bm{x}_1$) was generated as a sequence from $0$
to $1$ and length equals to the sample size.  Similarly, the categorical
covariate~($\bm{x}_2$) was generated as a sequence of three values each
one repeated $n/3$ times (rounding up when required), where $n$ denotes
the sample size. Thus, the expectation of the COM-Poisson random
variable is given by
$\bm{\mu} = \exp(\beta_0 + \beta_1 \bm{x}_1 + \beta_{21} \bm{x}_{21} +
\beta_{22} \bm{x}_{22})$,
where $\bm{x}_{21}$ and $\bm{x}_{22}$ are dummy representing the levels
of $\bm{x}_2$.  The regression coefficients were fixed at the values,
$\beta_0 = 2$, $\beta_1 = 05$, $\beta_{21} = 0.8$ and
$\beta_{22} = -0.8$.

We designed four simulation scenarios by considering four different
values of the dispersion parameter $\phi = -1.6, -1.0, 0.0$ and $1.8$.
Thus, we have strong and moderate overdispersion, equidispersion and
underdispersion, respectively.  \autoref{fig:justpars} shows the
variation of the average counts (left) and dispersion index (right) for
each value of the dispersion parameter considered in the simulation
study.  These configurations allow us to assess the properties of the
maximum likelihood estimators in extreme situations such as high counts
and low dispersion, and low counts and high dispersion, but also in the
standard case of equidispersion.

>>>>

% -------------------------------------------
% Abstract
\Abstract{ In the analysis of count data often the equidispersion
  assumption is not suitable, hence the Poisson regression model is
  inappropriate.  As a generalization of the Poisson distribution the
  COM-Poisson distribution can deal with under-, equi- and overdispersed
  count data.  It is a member of the exponential family of distributions
  and has the Poisson and geometric distributions as special cases, as
  well as the Bernoulli distribution as a limiting case.  In spite of
  the nice properties of the COM-Poisson distribution, its location
  parameter does not correspond to the expectation, which complicates
  the interpretation of regression models specified using this
  distribution. In this paper, we propose a straightforward
  reparametrization of the COM-Poisson distribution based on an
  approximation to the expectation of this distribution. The main
  advantage of our new parametrization is the straightforward
  interpretation of the regression coefficients in terms of the
  expectation of the count response variable, as usual in the context of
  generalized linear models. Furthermore, the estimation and inference
  for the new COM-Poisson regression model can be done based on the
  likelihood paradigm. We carried out simulation studies to verify the
  finite sample properties of the maximum likelihood estimators. The
  results from our simulation study show that the maximum likeli-hood
  estimators are unbiased and consistent for both regression and
  dispersion parameters. We observed that the empirical correlation
  between the regression and dispersion parameter estimators is close to
  zero, which suggests that these parameters are orthogonal. We
  illustrate the application of the proposed model through the analysis
  of three data sets with over-, under- and equidispersed count data.
  The study of distribution properties through a consideration of
  dispersion, zero-inflated and heavy tail indices, together with the
  results of data analysis show the flexibility over standard
  approaches. Therefore, we encourage the application of the new
  parametrization for the analysis of count data in the context of
  COM-Poisson regression models.  The computational routines for fitting
  the original and new version of the COM-Poisson regression model and
  the analyzed data sets are available in the supplementary material.
}

@@

Count data are random variables that assume non-negative integer values
and represent the number of times an event occurs in the observation
period. This kind of data is common in crop sciences, such as the number
of grains produced by a plant, number of fruits produced by a tree,
number of insects captured by a trap, to cite but a few.  Since the
seminal paper of \citet{Nelder1972} where the class of the generalized
linear models (GLMs) was introduced, the analysis of count data often
employs the Poisson regression model. This model provides a suitable
strategy for the analysis of count data and an efficient Newton scoring
algorithm can be used for fitting the model.

In spite of the advantages of the Poisson regression model, the Poisson
distribution has only one parameter, which represents both the
expectation and variance of the count random variable.  This restriction
on the relationship between the expectation and variance induced by the
Poisson distribution is referred as equidispersion.  However, in
practical data analysis such a restriction can be unsuitable, since the
observed data can present  variance both smaller or larger than the
expectation, leading to the cases of under and overdispersion,
respectively.  The main problem of the application of the Poisson
regression model to non-equidispersed count data is that the standard
errors associated with the regression coefficients are inconsistently
estimated, which in turn can lead to misleading
inferences~\citep{Winkelmann1995, Bonat2017}.

In practice, overdispersion is largely reported in the literature and
may occur due to the absence of relevant covariates, heterogeneity of
sampling units, different observational periods/regions not considered
in the analysis, and excess of zeros~\citep{Hinde1998}.  The case of
underdispersion is less report in the literature, however, it has been
of increasing interest in the statistical community. The processes that
reduce the variability are not as well-known as those leading to extra
variability.  For this reason, there are few approaches to deal with
underdispersed count data. The explanatory mechanisms leading to
underdispersion may be related to the underlying stochastic process
generating the count data.  When the time between events is not
exponentially distributed, the number of events can be over or
underdispersed, this process motivated the class of duration dependence
models~\citep{Winkelmann1995}.  Another possible explanation of
underdispersion is when the responses correspond to order statistics of
component observations, such as maxima of Poisson distributed
counts~\citep{Steutel1989}.

The strategies for constructing alternative count distributions are
related with the causes of the non-equidispersion. Specfically for the
overdispersion case Poisson mixture models are widely applied.  One
popular example of this approach is the negative-binomial model, where
the expectation of the Poisson distribution is assumed to be gamma
distributed.  However, other distributions can be used to represent the
random variation.  For example the Poisson-Tweedie
model~\citep{Bonat2017} and its special cases as the Poisson
inverse-Gaussian and Neyman-Type A assume that the random effects are
Tweedie, inverse Gaussian and Poisson distributed, respectively.  The
Gamma-Count distribution assumes a gamma distribution for the time
between events, thus it can deal with underdispersed as well as
overdispersed count data~\citep{Zeviani2014}. The COM-Poisson
distribution is obtained by a generalization of the Poisson distribution
allowing for a non-linear decrease in the ratios of successive
probabilities~\citep{Shmueli2005}.

The COM-Poisson distribution is a member of the exponential family and
it has the Poisson and geometric distributions as special cases, as well
as the Bernoulli distribution as a limiting case. It can deal with both
under and overdispersed count data.  Some recently applications of the
COM-Poisson distribution include \citet{Lord2010} for the analysis of
traffic crash data, \citet{Sellers2010} for the modelling of airfreight
breakage and book purchases, and \citet{Huang2017} to the analysis of
attendance data, takeover birds and cotton boll counts.  The main
disadvantage of the COM-Poisson regression model as presented
in~\citet{Sellers2010} is that its location parameter does not
correspond to the expectation of the distribution, which complicates the
interpretation of regression models and means that they are not
comparable with standard approaches such as the Poisson and negative
binomial regression models. \citet{Huang2017} proposed a
mean-parametrization of the COM-Poisson distribution in order to avoid
such an issue.  In this approach the mean parameter is obtained by
solving an non-linear equation defined as an infinite sum.
Consequently, it is computationally demanding and liable to numerical
problems.

The main goal of this article is to propose a novel COM-Poisson
parametrization based on the mean approximation presented by
\citet{Shmueli2005}. In this parametrization, the probability mass
function is written in terms of $\mu$ and $\phi$, where $\mu$ is the
expectation and $\phi$ is a dispersion parameter. In contrast to the
original parametrization, the proposed parametrization leads to
regression coefficients directly associated with the expectation of the
response variable, as usual in the context of generalized linear models.
Consequently, the obtained regression coefficients are comparable with
the ones obtained by standard approaches, such as the Poisson and
negative binomial regression models.  Furthermore, our novel COM-Poisson
parametrization is simpler than the strategy proposed
by~\cite{Huang2017}, since it does not require any numerical method for
solving non-linear equations, and we show the attractive properties like
the orthogonality between dispersion and regression parameters and
consistency and asymptotic normality of the maximum likelihood
estimators are retained.

This paper is organized as follows. In \autoref{background} we present
the COM-Poisson distribution and the strategy proposed
by~\cite{Huang2017}.  The proposed reparametrization, assessment of
moment approximations, and study of distribution properties are
considered in the~\autoref{reparametrization}. In
the~\autoref{estimation-and-inference} we present estimation and
inference for the novel COM-Poisson regression model based on the
likelihood paradigm. The properties of the maximum likelihood estimators
and the orthogonality property are assessed
in~\autoref{simulation-study} through simulation studies.  We illustrate
the application of the new COM-Poisson regression model through the
analysis of three data sets. We provide an \texttt{R} implementation of
the COM-Poisson and reparameterized COM-Poisson regression models as
well as the analyzed data sets in the supplementary
material.\footnote{Available on
  \texttt{\url{http://www.leg.ufpr.br/~eduardojr/papercompanions}}
  \label{papercompanion}.}.

@@

The COM-Poisson distribution generalizes the Poisson distribution in
terms of the ratio between the probabilities of two consecutive events
by adding an extra dispersion parameter~\citep{Sellers2010}.  Let $Y$ be
a COM-Poisson random variable, then
$$\frac{\Pr(Y=y-1)}{\Pr(Y=y)} = \frac{y^\nu}{\lambda}$$ while for the
Poisson distribution this ratio is $\frac{y}{\lambda}$ corresponding to
$\nu=1$.  It allows the COM-Poisson distribution deals with
non-equidispersed count data.  The probability mass function of the
COM-Poisson distribution is given by
\begin{equation}
  \label{eqn:pmf-cmp}
  \Pr(Y=y \mid \lambda, \nu) = \frac{\lambda^y}{(y!)^\nu Z(\lambda, \nu)},
  \qquad y = 0, 1, 2, \ldots,
\end{equation}
where $\lambda > 0$, $\nu \geq 0$ and $Z(\lambda, \nu) =
\sum_{j=0}^\infty \frac{\lambda^j}{(j!)^\nu}$ is a normalizing
constant that depends on both parameters.

The $Z(\lambda, \nu)$ series diverges theoretically only when $\nu=0$
and $\lambda \geq 1$, but numerically for small values of $\nu$ combined
with large values of $\lambda$, the sum is so huge it causes
overflow. \autoref{tab:convergenceZ} shows the values of the normalizing
constants using one thousand increments, that is,
$\sum_{j=0}^{1000}\lambda^j/(j!)^\nu$ for different values of $\lambda$
and $\phi$.

@@

In the first line of \autoref{tab:convergenceZ} we have mathematically
divergent series, because $\sum_{j=0}^\infty\lambda^j$ is divergent when
$\lambda \geq 1$.  In other cases the series diverges numerically, due
to the computational storage limitation.  For both forms of divergence
it is impossible to compute probabilities, therefore, this acts as a
restriction on the parameter space.

@@

The COM-Poisson regression model was proposed by~\citet{Sellers2010},
using the original parametrization. In this case, the COM-Poisson
regression model is $\log(\lambda_i) = \bm{x}_i^\top \bm{\beta}$ and the
relationship between E$(Y_i)$ and $\bm{x}_i$ is modelled indirectly.
\citet{Huang2017} shows how to model directly the expectation of the
COM-Poisson distribution in a suitable parametrization. In the
\autoref{eqn:pmf-cmp}, \Citeauthor{Huang2017} proposes that the
parameter $\lambda$ as a function of $\mu$ and $\nu$, is given by the
solution to
\begin{equation*}
  \sum_{j=0}^{\infty} (j - \mu) \frac{\lambda^j}{(y!)^\nu} = 0\,.
\end{equation*}
Thus the mean-parametrized COM-Poisson regression model is $\log(\mu_i)
= \bm{x}_i^\top \bm{\beta}$. In this article, we propose an alternative
mean-parametrization of the COM-Poisson distribution in order to avoid the
limitations of the original parametrization and the numerical complexity
of the \Citeauthor{Huang2017}'s approach.

@@

The proposed reparametrization of COM-Poisson models is based on
the mean approximation (\autoref{eqn:mean-approx}). We introduced a new
parameter $\mu$, using this approximation,
\begin{equation}
  \label{eqn:repar-cmp}
  \mu = h(\lambda, \nu) = \lambda^{1/\nu} - \frac{\nu - 1}{2\nu}
  \quad \Rightarrow \quad
  \lambda = h^{-1}(\mu, \nu) = \left (\mu +
    \frac{(\nu - 1)}{2\nu} \right )^\nu.
\end{equation}
The dispersion parameter is taken on the log scale for computational
convenience, thus $\phi = \log(\nu)$, $\phi \in \mathbb{R}$. The
interpretation of $\phi$ is the same as the $\nu$, but on another
scale. For $\phi < 0$ and $\phi > 0$ we have the overdispersed and
underdispersed cases, respectively.  When $\phi=0$ we have Poisson
distribution as a special case.

@@

In order to assess the accuracy of the moment approximations
(\autoref{eqn:mean-approx}), \autoref{fig:approx-plot} presents the
quadratic errors for (a) expectation  and (b) variance.  The
quadratic errors were obtained by $[\mu - \text{E}(Y)]^2$ for the
expectation and by $[ \mu \exp(-\phi) - \text{Var}(Y)]^2$ for the
variance. In both cases $\text{E}(Y)$ and Var$(Y)$ were computed
numerically. The dotted lines represent the border between the regions
$\nu \leq 1$ and $\lambda > 10^\nu$, in the $\mu$ and $\phi$ scale.

<<approx-plot, fig.height=3, fig.width=6, out.width=".8\\textwidth", fig.cap="Quadratic errors for the approximation of the (a) expectation and (b) variance. Dotted lines represent the restriction for suitable approximations given by \\cite{Shmueli2005}.">>=

@@

The results in \autoref{fig:approx-plot} show that the mean
approximation is accurate, the largest quadratic error is
\Sexpr{max(grid[["emu"]])} for the parameter values evaluated. For the
variance approximation, the largest quadratic error was
\Sexpr{max(grid[["eva"]])} and it occurs for negative values of
$\phi$. Interestingly, the errors are larger for negative values of
$\phi$ and present no clear relation with $\mu$, as opposed to the
regions gives by \citet{Shmueli2005} ($\phi \leq 0$ and
$\mu > 10 - \frac{\exp(\phi) - 1}{2\exp(\phi)}$).

@@

These indexes are discussed by \citet{Bonat2017} to study the
flexibility of Poisson-Tweedie distribution, and \citet{Puig2006} to
describe count distributions. Regarding the COM-Poisson$_\mu$
distribution, in \autoref{fig:indexes-plot} we present the relationship
between (a) mean and variance, (b--c) the dispersion and zero-inflation
indexes for different values of $\mu$ and $\phi$, and (d) heavy-tail
index for $\mu=25$ and different values of $y$ and $\phi$.

@@

<<indexes-plot, fig.width=9.5, fig.height=4, fig.cap="Indexes for COM-Poisson distribution. (a) Mean and variance relationship, (b--d) dispersion, zero-inflation and heavy-tail indexes for different parameter values. Dotted lines represents the Poisson special case.">>=

@@

\autoref{fig:indexes-plot} shows that the indexes are slightly dependent
on the expected values and tend to stabilize for large values of
$\mu$. Consequently, the mean and variance relationship
\autoref{fig:indexes-plot}(a) is proportional to the dispersion
parameter $\phi$. In terms of moments, this leads to a specification
indistinguishable from the quasi-Poisson regression model. The
dispersion indexes in \autoref{fig:indexes-plot}(b) show that the
distribution is suitable to deal to dispersed counts, of course. For the
parameter values evaluated the largest DI was \Sexpr{madi} and smallest
was \Sexpr{midi}. \autoref{fig:indexes-plot}(c) shows the COM-Poisson
can handle a limited amount of zero-inflation, in cases of
overdispersion ($\phi < 0$). On the other hand, for $\phi > 0$
(underdispersion) this distribution is suitable to deal with
zero-deflated counts. Heavy-tail indexes in
\autoref{fig:indexes-plot}(d) indicate the distribution is in general a
light-tailed distribution, i.e. $HT \to 0$ for $y \to \infty$.

@@

In this section we describe the estimation and inference for the
two forms of the COM-Poisson regression model based on the maximum
likelihood method. The log-likelihood function for a set of independent
observations $y_i$, $i=1,2,\ldots,n$ from the COM-Poisson$_\mu$
distribution has the following form,
\begin{equation}
  \label{eqn:ll-rcmp}
  \ell = \ell(\bm{\beta}, \phi \mid \bm{y}) =
  e^\phi \left [
    \sum_{i=1}^n y_i
    \log \left( \mu_i + \frac{e^\phi-1}{2e^\phi} \right ) -
    \sum_{i=1}^n \log(y_i!) \right ] -
  \sum_{i=1}^n \log(Z(\mu_i, \phi)),
\end{equation}
where $\mu_i = \exp(\bm{x}_i^\top\bm{\beta})$, with
$\bm{x}_i^\top = (x_{i1},\, x_{i2},\, \ldots,\, x_{ip})$ is a vector of
known covariates for the $i$-th observation, and $(\bm{\beta},\, \phi)
\in \mathbb{R}^{p+1}$. The normalizing constant $Z(\mu_i, \phi)$ is
given by
\begin{equation}
\label{eqn:infseries}
  Z(\mu_i, \phi) = \sum_{j=0}^\infty \left [ \left (
    \mu_i + \frac{e^\phi - 1}{2e^\phi} \right )^{je^\phi}
  \frac{1}{(j!)^{e^\phi}} \right ].
\end{equation}
The evaluation of the log-likelihood function for each observation
involves the computation of the infinite series
(\autoref{eqn:infseries}).  Thus, the fitting procedure is computationally
expensive for regions of the parameter space where the convergence of
the infinite sum is slow.

@@

In the applications we fitted the quasi-Poisson model
\citep{Wedderburn1974} as a baseline model. This approach is based on a
second-moment assumption that allows more flexibility to the model. In
this case the variance of the response variable is fixed by an
additional parameter $\sigma$, $\textrm{Var}(Y_i)=\sigma \mu_i$. These
models are fitted in the \texttt{R} software using the function
\texttt{glm(..., family = quasipoisson)}.

@@

We considered average counts varying from $3$ to $27$ according to a
regression model with a continuous and a categorical covariate. The
continuous covariate~($\bm{x}_1$) was generated as a sequence from $0$
to $1$ and of length equal to the sample size.  Similarly, the categorical
covariate~($\bm{x}_2$) was generated as a sequence of three values each
one repeated $n/3$ times (rounding up when required), where $n$ denotes
the sample size. Thus, the expectation of the COM-Poisson random
variable is given by
$\bm{\mu} = \exp(\beta_0 + \beta_1 \bm{x}_1 + \beta_{21} \bm{x}_{21} +
\beta_{22} \bm{x}_{22})$,
where $\bm{x}_{21}$ and $\bm{x}_{22}$ are dummy representing the levels
of $\bm{x}_2$.  The regression coefficients were fixed at the values,
$\beta_0 = 2$, $\beta_1 = 0.5$, $\beta_{21} = 0.8$ and
$\beta_{22} = -0.8$.

We designed four simulation scenarios by considering different
values of the dispersion parameter $\phi = -1.6, -1.0, 0.0$ and $1.8$.
Thus, we have strong and moderate overdispersion, equidispersion, and
underdispersion, respectively.  \autoref{fig:justpars} shows the
variation of the average counts (left) and dispersion index (right) for
each value of the dispersion parameter considered in the simulation
study.  These configurations allow us to assess the properties of the
maximum likelihood estimators in extreme situations, such as high counts
and low dispersion, and low counts and high dispersion, but also in the
standard case of equidispersion.
